---
title: "Introduction Problem Set"
author: "insert your name here"
date: "insert date here"
output: pdf_document
classoption: landscape

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to Data Management and Manipulation Using R!

In this problem set, you will practice "knitting," loading data from a url, and loading data from absolute/relative filepaths.

- Note: Change author to your name & date to current date above in the YAML header (we'll learn more about headers later on...)
- If you're experiencing errors while working on this assignment (or any assignment), start fresh by closing out R (when prompted to "save your workspace" hit "don't save) and re-opening the assignment. 
- You may also be experiencing errors for this particular assignment because of running R chunks for questions below out of order (run Question 1, then Question 2, then Question 3) 

\vspace{30pt}

### Step 1: Practice "knitting" the problem set Rmd file!

- Open your RStudio via the Rproject (`HED96C_RClass.Rproj`) we created in class
- Once in RStudio, in the top menu bar, click on File > Open File...
- Navigate to your HED696C_RClass folder and open your saved `introduction_ps.Rmd` file (hint it should be located in: problemsets/introduction folder)
- Open the introduction_ps.Rmd file in RStudio
- At the top of the introduction_ps.Rmd file, insert your name and the date in the first few lines of this .Rmd file where indicated for you
- Now select the "Knit" tab (icon with blue yarn ball) or use the drop down menu next to the yard ball and select "Knit to PDF"
- The goal of this step is to make sure you are able to knit to a PDF without running into errors. But it is OK if you do! We'll figure them out together!

\vspace{30pt}

### Step 2: Practice with Directories! 

**Question 1: Load .Rdata directly with url** 

- In the R Chunk Below....
- View current/working directory first using the `getwd()` function
  - The current working directory when running from an R Chunk will revert to the location of this introduction_ps.Rmd file; which should be in the problemsets folder within the introduction sub-folder
  - If ever you don't know what a "function" does; you can ask R directly using the syntax `?` before the function. 
- We're going to open a dataset on my "recruiting" project which is stored in an online repository. 
  - __Here is the link to that dataset:__
  - https://github.com/ksalazar3/HED696C_RClass/blob/master/data/recruiting/recruit_ps1_small.Rdata?raw=true
- When the data is stored online/in a repository, we can load R datasets (their file extensions are .Rdata) directly into R using the `load()` + `url() ` functions, or `load(url())` when combined,  by using the link! 
- So the "generic" syntax will look like this: 
  - `load(url("INSERT THE LINK HERE"))`
- Whereas the filled in syntax for this part of the question will look like this:
  - `load(url("https://github.com/ksalazar3/HED696C_RClass/blob/master/data/recruiting/recruit_ps1_small.Rdata?raw=true"))`

_Solutions provided; see how I followed the instructions to get the syntax answers in the R Chuck below_
```{r}
getwd()
?getwd()
?load
load(url("https://github.com/ksalazar3/HED696C_RClass/blob/master/data/recruiting/recruit_ps1_small.Rdata?raw=true"))
```


\vspace{30pt}


**Question 2: Set your working directory using an absolute filepath; then load the data directly from your computer**  

- Download .Rdata using link from Question 1 above
- Move the `recruit_ps1_small.Rdata` file (should be in your downloads folder) in the recruiting data folder (HED696C_RClass/data/recruiting)
- Then, in the R Chunk Below....
- View your current/working directory `getwd()` first
  - This will revert to the location of the introduction_ps.Rmd file; which should be in the problemsets folder within the introduction sub-folder
- Then use the `setwd()` function to set your working directory to where the .Rdata file is saved using the **absolute filepath** 
  - Hint: This should be pointing to the recruiting folder where the .Rdate file is located, it should not include the name of the Rdata file. 
  - For example, my absolute file path is "/Users/karinasalazar/Dropbox/HED696C_RClass/data/recruiting"
  - So MY solution for this part of the question would be
     `setwd("/Users/karinasalazar/Dropbox/HED696C_RClass/data/recruiting")`
  - After setting your working directory you can use `getwd()` again to see how they changed!
- After setting the your working directory using the absolute filepath, load the .Rdata file by using the following generic syntax `load(INSERT NAME OF DATAFILE HERE.Rdata)`
- Using the `names(data frame name)` function, show the column names in this dataset  (Hint: data frame name for the `recruit_ps1_small.Rdata` is called df_recruiting)  


_Write your solutions in the R Chunk below_
```{r}

```

\vspace{30pt}


**Question 3**: Set your working directory using a relative filepath; then load the data directly from your computer** 

- In the R Chunk below... 
- View your current/working directory `getwd()`
  - This will revert to the location of the introduction_ps.Rmd file; which should be in the problemsets folder within the introduction sub-folder
- Then use the `setwd()` function to set your working directory to where the .Rdata file is saved using the **relative filepath** 
  - Hint: you are currently in the problemsets/introduction folder; so you will need to move "up" or "back" before you can get to the data/recruiting folder
  - For example, if my current working directory is:     "/Users/karinasalazar/Dropbox/HED696C_RClass/problemsets/introduction"
  - So my solution for this part of the question would be
     `setwd("../../data/recruiting")`
- After setting the your working directory using the absolute filepath, load the .Rdata file by using the following generic syntax `load(INSERT NAME OF DATAFILE HERE.Rdata)`
- Show the column names in this dataset (Hint: data frome name is df_recruiting)  

_Write your solutions in the R Chunk below_
```{r}

```



\vspace{30pt}


**Question 4**: Without first changing the working directory, load the dataset using relative filepath 

- View your current/working directory `getwd()`
- Load data without setting directory `load("relativefilepath/recruit_ps1_small.Rdata")`
- Hint: this will look like a combination of the two steps in Question 3


_Write your solutions in the R Chunk below_
```{r}

```

\vspace{30pt}

**Knit to pdf** select the "Knit" tab (icon with blue yarn ball) or scroll down and "Knit to PDF"  
- Submit both .Rmd and pdf
- Use this naming convention "lastname_introduction_ps.Rmd"
