---
title: "Problem Set 1"
author: "insert your name here"
date: "insert date here"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to Data Management and Manipulation Using R!

In this problem set, you will load data from a url, or absolute/relative filepath.

- Note: Change author to your name & date to current date above in the YAML header (we'll learn more about headers later on...)


### Step 1: Practice "knitting" the lecture files!

- Open your RStudio via the Rproject (`HED96C_RClass.Rproj`) we created in class
- Once in RStudio, in the top menu bar, click on File > Open File...
- Navigate to your HED696C_RClass folder and open your saved `introduction.Rmd` file (hint it should be located in: modules/introduction folder)
- Open the introduction.Rmd file in RStudio
- At the top of the introduction.Rmd file, change my name under "author" to your first and last name
- Now select the "Knit" tab (icon with blue yarn ball) or use the drop down menu next to the yard ball and select "Knit to PDF"
- The goal of this step is to make sure you are able to knit lecture slides to a PDF without running into errors. But it is OK if you do! We'll figure them out together!


### Step 2: Practice with Directories! 

**Question 1**: Load .Rdata directly with url  

- View current/working directory first using the `getwd()` function
  - If ever you don't know what a "function" does; you can ask R directly using the syntax `?` before the function. See my example below in the R chunk
- We're going to open a dataset on my "recruiting" project which is stored in an online repository. Here is the link to that dataset: https://github.com/ksalazar3/HED696C_RClass/blob/master/data/recruiting/recruit_ps1_small.Rdata  
- When the data is stored online/in a repository, we can load R datasets (they file extensions are .Rdata) directly into R using the `load()` and `url() ` functions by using the link! 
  - There's no need to download the file if we open the data directly from a link!
- So the syntax will look like this: `load(url("link"))`  

```{r}
?getwd()
getwd()
?load
load(url("https://github.com/ksalazar3/HED696C_RClass/raw/master/data/recruiting/recruit_ps1_small.Rdata"))
```


**Question 2**: Set your working directory and use absolute filepath to load data  

- Download .Rdata using link from above  
- Save .Rdata file in the recruiting folder (rclass/data/recruiting)  
- View your current/working directory `getwd()` 
- Use `setwd()` function to set your working directory to where .Rdata file is saved (rclass/data/recruiting)  
- When setting working directory, use the absolute filepath  
- Load the .Rdata file  `load(name.Rdata)`
- Using the `names(data frame name)` function, report the names in this dataset  (Hint: data frame name is df_recruiting)  

```{r}
getwd() 
setwd("/Users/karinasalazar/Desktop/rclass/data/recruiting") #absolute file path to data
getwd()  #if you run lines entire chunk so you can see how the working directory changes
list.files() #list.files displays the files in the folder
load("recruit_ps1_small.Rdata")
names(df_recruiting)
```

**Question 3**: Set your working directory and use relative filepath to load data  

- View your current/working directory `getwd()` 
- Use `setwd()` function to set your working directory to where .Rdata file is saved (rclass/data/recruiting)  
- When setting working directory, use the relative filepath  
- Load the .Rdata file  `load()`
- Report the names in this dataset (Hint: data frome name is df_recruiting)  

```{r}
getwd() 
# Working directory is in the lecture1 folder, need to get to recruiting folder using relative file path
setwd("../../data/recruiting")
getwd() #View current directory after setting the directory
list.files() #listing files in the recruiting folder
# Lets load recruit_ps1_small.Rdata 
load("recruit_ps1_small.Rdata")
# Use names function to view the names in the data set (data frame)
names(df_recruiting)
# Run the entire code chunck 
```

**Question 4**: Load dataset using relative filepath without first changing working directory  

- View your current/working directory `getwd()`  
- Load data without setting directory `load(filepath/data.Rdata)`

```{r}
getwd()
load("../../data/recruiting/recruit_ps1_small.Rdata") #Set file path and load data at once
names(df_recruiting)
```

**Knit to pdf** select the "Knit" tab (icon with blue yarn ball) or scroll down and "Knit to PDF"  

- Submit both .Rmd and pdf   
- Use this naming convention "lastname_firstname_ps#"
