---
title: Managing and Manipulating Data Using R # potentially push to header
subtitle:  "Module 1: Objects in R"
author:
date: 
urlcolor: blue
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: # toc_float option to float the table of contents to the left of the main document content. floating table of contents will always be visible even when the document is scrolled
      collapsed: false # collapsed (defaults to TRUE) controls whether the TOC appears with only the top-level (e.g., H2) headers. If collapsed initially, the TOC is automatically expanded inline when necessary
      smooth_scroll: true # smooth_scroll (defaults to TRUE) controls whether page scrolls are animated when TOC items are navigated to via mouse clicks
    number_sections: TRUE
    fig_caption: true # ? this option doesn't seem to be working for figure inserted below outside of r code chunk    
    highlight: default # Supported styles include "default", "tango", "pygments", "kate", "monochrome", "espresso", "zenburn", and "haddock" (specify null to prevent syntax    
    theme: default # theme specifies the Bootstrap theme to use for the page. Valid themes include default, cerulean, journal, flatly, readable, spacelab, united, cosmo, lumen, paper, sandstone, simplex, and yeti.
    df_print: tibble #options: default, tibble, paged
    keep_md: true # may be helpful for storing on github
    
---

```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
  #comment = "#>" makes it so results from a code chunk start with "#>"; default is "##"
```

### As you wait for class to begin... {-}

- Download the module1.Rmd lecture file from the class website [https://ksalazar3.github.io/HED696C_RClass/](https://ksalazar3.github.io/HED696C_RClass/)
- Move the module1.Rmd file to "HED696C_RClass/modules/module1"
- Open RStudio via the `HED696C_Rclass.rproj`
- Once in R Studio go to File > Open File > navigate to and open the module1.Rmd
- Try knitting the module1.Rmd file (it will open an HTML document)

# Problem Set Review

### Working Directories {-}

- Your automatic working directory will change based on *how* you are using R...
- There are some basic hierarchies....
  (1) Rmd files
  (2) R Projects
  (3) R Scripts
- But overall lots of nuances... 
  - You don't have to remember these...
  - You do want to begin developing some general intuition in how R works with directories
- You should develop a skill set to *figure out* how to check working directories and moving back/forth no matter the nuances of how R works with directories
  - `getwd()` 
  - `setwd()`
  - absolute versus relative filepaths
  

### Working Directories via R Projects {-}

- An R project will open up your R Studio to the working directory you specified when you created the .rproj file... (like weekly lectures)
- When you open an __R project__, your current working directory is automatically set to where your .rproj file is located
  - If you're working in an __R script__, your working directory stays where the .rproj file is located
  - If you're working in an __Rmd file__, your working directory will revert to where the .rmd file is located
- Show in R....
```{r}
getwd()
```

 
### Working Directories via R Scripts {-}

- If you've opened R Studio via an R project...
  - The R script will begin by automatically setting your current working directory to where your .rproj file is located
  - From there, the working directory will only change if YOU change it
- If you did NOT open R Studio via an R project...
  - The R script will begin by automatically setting your current working directory to your "root" user

 

### Working Directories via Rmd Files {-}

- When working within an Rmd file... (like your weekly problem sets)
  - Your working directory will be automatically set to the folder where the Rmd file is located [even if you are working out of an R project]
  - When you change your working directory within an R chunk, via `setwd()`, the new working directory will only be set for that one particular chunk [e.g., warning message]
  - When you move on to another R chunk in the same Rmd file, your working directory will reset back to where the Rmd file is located

Changing working directory, first chunck
```{r}
getwd()
setwd("../../data/recruiting")
getwd()
```


Changing working directory, second chunck
```{r}
getwd()
```

  

# R basics
  

### R as a calculator

```{r}
5
5+2
10*3
```

### Executing commands in R

```{r}
5
5+2
10*3
```

Three ways to execute commands in R

1. Type/copy commands directly into the "console"
1. `code chunks' in RMarkdown (.Rmd files)
    - Can execute one command at a time, one chunk at a time, or "knit" the entire document
1. R scripts (.R files)
    - This is just a text file full of R commands
    - Can execute one command at a time, several commands at a time, or the entire script

### Shortcuts you should learn for executing commands

```{r}
5+2
10*3
```

Three ways to execute commands in R

1. Type/copy commands directly into the "console"
1. `code chunks' in RMarkdown (.Rmd files)
    - __Cmd/Ctrl + Enter__: execute highlighted line(s) within chunk
    - __Cmd/Ctrl + Shift + k__: "knit" entire document
1. R scripts (.R files)
    - __Cmd/Ctrl + Enter__: execute highlighted line(s)
    - __Cmd/Ctrl + Shift + Enter__ (without highlighting any lines): run entire script


### Comments in R

- Once you start writing significant code (either in an R script or within R chunks inside of a Rmd file), you will need to make comments on what the code is/doing
- Comments need to be treated as "special" characters so that R does not interpret it as "code/syntax"
  - We specify comments via `#` either at beginning of a line OR after code/syntax on the same line
  - When working in an Rchunk anything the console prints will begin with `#>`
- Show an R script example...

```{r}
# This is a comment; the code/syntax below is not a comment
getwd()
?getwd # This is also a comment; this code pulls the help file for the function

```



### Assignment

__Assignment__ means creating a variable -- or more generally, an "object" -- and assigning values to it

- `<-` is the assignment operator
    - in other languages `=` is the assignment operator
- good practice to put a space before and after assignment operator

```{r}
# Create an object and assign value
a <- 5

# Printing the `a` object I just created by simply "calling" it
a

b <- "yay!"

# Printing the `b` object I just created by simply "calling" it
b
```

# Classification of objects

### Some Note on Objects... {-}

- The remainder of the lecture is a conceptual and practical introduction to "objects" in R
- This lecture is a broad overview that introduces some general programming and applies some of these concepts to data
- __IMPORTANT__: The goal is to BEGIN developing familiarity with the concepts that will be introduced in more detail in later weeks.
  - I don't expect you to understand or retain all this information perfectly
  - Just focus on trying to understand as much as you can...


### Objects {-}

Most statistical software (e.g., SPSS, Stata) operates on datasets, which consist of rows of observations and columns of variables

- Usually, these packages can open only one dataset at a time

R is an "object-oriented" programming language (like Python, JavaScript). So, what is an "object"?

- Formal computer science definitions are confusing because they require knowledge of concepts I haven't introduced yet
- More intuitively, I think objects as anything I assign values to
    - For example, below, "`a`" and "`b`" are objects I assigned values to
```{r}
a <- 5
a
b <- "yay!"
b
```
    

- "Objects are like boxes in which we can put things: data, functions, and even other objects."

### Object Type and Class {-}

- Objects can be categorized by "__type__" (which we will discuss today) and by "__class__" (which we will discuss in later weeks)
    - e.g., a date is an object with a numeric _type_ and a date _class_
- There is no limit to the number of objects R can hold (except memory)
- R "functions" (e.g., `setwd()`, `getwd()`) do different things to different types/classes of objects
  - e.g., date functions (e.g., `as.Date()`) are meant to process objects with _type_=numeric and _class_=date; these functions don't work on objects with _type_=character (e.g., "yay!")


### Vectors {-}

The fundamental object in R is the "vector"

- A vector is a collection of values
- The individual values within a vector are called "elements"
- Values in a vector can be numeric, character (e.g., "Apple"), or some other _type_

Below I use the combine function `c()` to create a numeric vector that contains three elements

- Help file says that `c()` "combines values into a vector"

`x` is a Vector where the elements are numeric
```{r}
?c
x <- c(4, 7, 9)
x
```

`animals` is a Vector where the elements are characters
```{r}
animals <- c("lions", "tigers", "bears", "oh my")
animals
```


### Student task (10 minutes on your own) {-}

__Open a new R script for module1:__

1. In R Studio > File (top navigation bar) > New File > R Script
1. In the new R Script, write a comment that titles the R script for module 1 (ex: `# Module 1 R Script `)
1. Save this file as `module1.R` in the "HED696C_RClass/modules/module1" folder

\medskip

__In `module1.R`:__

1. Create a vector called `v1` with three elements, where all the elements are numbers. Then print the values.
1. Create a vector called `v2` with four elements, where all the elements are characters (i.e., enclosed in single '' or double "" quotes). Then print the values.
1. Create a vector called `v3` with five elements, where some elements are numeric and some elements are characters. Then print the values.

### Formal classification of vectors in R {-}

Here, I introduce the classification of vectors by Wickham and Grolemund

There are two broad types of vectors

1. __Atomic vectors__. An object that contains elements. 
    - Atomic vector elements are all the same "type"
    - There are six types of atomic vectors based on the type of elements they hold: 
    - __logical__, __integer__, __double__, __character__, __complex__, and __raw__. 
        - __Integer__ and __double__ vectors are collectively known as __numeric__ vectors.

2. __Lists__. Like atomic vectors, lists are objects that contain elements
    - But elements in lists can be different
    - Elements within a list may be atomic vectors
    - Elements within a list may also be other lists; that is lists can contain other lists
    - This sounds vague and confusing; I'll give examples below
    
One difference between atomic vectors and lists: __homogeneous__ vs. __heterogeneous__ elements

* Atomic vectors are __homogeneous__: all elements within atomic vector must be of the same type
* Lists can be __heterogeneous__: e.g., one element can be an integer and another element can be character



<!-- ### Developing an intuitive understanding of vector types {-} -->

<!-- __Grolemund and Wickham classification__: -->

<!-- 1. __Atomic vectors__. six types: logical, integer, double, character, complex, raw.  -->
<!-- 2. __Lists__ -->

<!-- Problem with this classification: -->

<!-- - Not conceptually intuitive -->
<!-- - Technically, lists are a type of vector, but people often think of atomic vectors and lists as fundamentally different things -->

<!-- __Classification that I use__: -->

<!-- - data __type__: logical, numeric (integer and double), character, etc. -->
<!-- - data __structure__: vector, list, matrix, etc. -->

<!-- I find Skinner's classification more intuitive conceptually. However, it isn't consistent with R functions or the way R thinks about objects.  -->

<!-- If you find this classification of data _type_ and data _structure_ helpful, totally fine to think of objects in this way while you start to learn R. -->

# Atomic vectors

### "Length" of an atomic vector is the number of elements {-}

\medskip For remainder of lecture, I'll use the term __vector__ to refer to atomic vectors

Use `length()` function to examine vector length
```{r}
x <- c(4, 7, 9)
x
length(x)

animals <- c("lions", "tigers", "bears", "oh my")
animals
length(animals)
```

A single number (or a single string/character) is a vector with `length==1`
```{r}
z <- 5
length(z)
length("Tommy")
```


### Data type of a vector {-}

\medskip The "type" of an atomic vector refers to the elements within the vector. 

While there are six "types" of actomic vectors, we'll focus on the following types:

- numeric: 
    - "integer" (e.g., 5)
    - "double" (e.g., 5.5)
- character (e.g., "karina")
- logical (e.g., `TRUE`, `FALSE`)

Use `typeof()` function to examine vector type
```{r}
x
typeof(x)

p <- c(1.5, 1.6)
p
typeof(p)

animals
typeof(animals)
```

### Data type of a vector, numeric {-}

Numeric vectors can be "integer" (e.g., 5) or "double" (e.g., 5.5)


```{r}
typeof(1.5)
```

R stores numbers as doubles by default.
```{r}
x
typeof(x)

```

To make an integer, place an `L` after the number:
```{r}
typeof(5)
typeof(5L)
```

### Data type of a vector, character {-}

In contrast to "numeric" data types which are used to store numbers, the "character" data type is used to store __strings__ of text.  

- Strings may contain any combination of numbers, letters, symbols, etc.
- Character vectors are sometimes referred to as string vectors

When creating a vector where elements have `type==character` (or when referring to the value of a string), place single '' or double "" quotes around text

- the text within quotes is the "string"

```{r}
c1 <- c("cat",'cash','candy cane')
c1
typeof(c1)
length(c1)
```
Numeric values can also be stored as strings
```{r}
c2 <- c("1","2","3")
c2
typeof(c2)
```


### Data type of a vector, logical {-}

Logical vectors can take three possible values: `TRUE`, `FALSE`, `NA`

- `TRUE`, `FALSE`, `NA` are special keywords; they are different from the character strings `"TRUE"`, `"FALSE"`, `"NA"`
- Don't worry about `"NA"` for now

```{r}
typeof(TRUE)
typeof("TRUE")

typeof(c(TRUE,FALSE,NA))
typeof(c(TRUE,FALSE,NA,"FALSE"))

log <- c(TRUE,TRUE,FALSE,NA,FALSE)
typeof(log)
length(log)
```
We'll learn more about logical vectors later

### All elements in (atomic) vector must have same data type {-}

Atomic vectors are __homogenous__;

- An atomic vector has one data type
- all elements within an atomic vector must have the same data "type"

If a vector contains elements of different type, the vector type will be type of the most "complex" element

Atomic vector types from simplest to most complex: 

- logical < integer < double < character

```{r}
typeof(c(TRUE,TRUE,NA))
typeof(c(TRUE,TRUE,NA,1L)) # recall L after an integer forces type to be integer rather than double
typeof(c(TRUE,TRUE,NA,1.5))
typeof(c(TRUE,TRUE,NA,1.5,"howdy!"))
```

### Named vectors {-}

All vectors can be "named" (i.e., name individual elements within vector)
  
  - hint: dataframes with variable names

Example of creating an unamed vector

- the `str()` function "compactly display[s] the internal structure of an R object" [from help file]; very useful for describing objects
```{r}
#?str
x <- c(1,2,3,"hi!")
x
str(x)
```

Example of creating a named vector
```{r}
y <- c(a=1,b=2,3,c="hi!")
y
str(y)
```
### Sequences {-}

(Loose) definition: a sequence is a set of numbers in ascending or descending order

A vector containing a "sequence" of numbers (e.g., 1, 2, 3) can be created using the colon operator `:` with the notation `start:end`
```{r}
-5:5
5:-5
s<- 1:10 #same as this: s<- c(1:10)
s
length(s)
```
Creating sequences using `seq()` function

- basic syntax [with default values]: 
```{r, eval=FALSE}
seq(from = 1, to = 1, by = 1)
```

```{r}
seq(10,15)
seq(from=10,to=15,by=1)
seq(from=100,to=150,by=10)
```

### Vectorized math {-}

Most mathematical operations operate on each element of the vector

- e.g., add a single value to a vector and that value will be added to each element of the vector
```{r}
1:3
1:3+.5
(1:3)*2
```

- e.g., for addition: add element 1 of vector 1 to element 1 of vector 2, add element 2 of vector 1 to element 2 of vector 2, etc.
```{r}
c(1,1,1)+c(1,0,2)
c(1,1,1)*c(1,0,2)
```

# Lists

### Lists {-}

What is a __list__?

- Like (atomic) vectors, a list is an object that contains __elements__
- Unlike vectors, data types can differ across elements within a list
- An element within a list can be another list
    - this characteristic makes lists more complicated than vectors
    - suitable for representing hierarchical data

Lists are more complicated than vectors; today I'll just provide a basic introduction

### Create lists using `list()` function {-}

Create a vector (for comparison purposes)
```{r}
a <- c(1,2,3)
typeof(a)
length(a)
```
Create a list
```{r}
b <- list(1,2,3)
typeof(b)
length(b)
b # print list is awkward
```


### Investigate structure of lists using `str()` function {-}

When investigating lists, `str()` is better than printing the list

```{r}
b <- list(1,2,3)
typeof(b)
length(b)
str(b) # 3 elements, each element is a numeric vector w/ length=1
```

```{r}
c <- list(c(3,4),c(-5,1,3))
typeof(c)
length(c)
str(c) # 2 elements; element 1=vector w/ length=2; element 2=vector w/length=3
```

### Elements within lists can have different data types {-}

Lists are __heterogeneous__

- data types can differ across elements within a list
```{r}
b <- list(1,2,"apple")
typeof(b)
length(b)
str(b)
```
Vectors are __homogeneous__
```{r}
a <- c(1,2,"apple")
typeof(a)
str(a)
```

### Lists can contain other lists {-}

```{r}
x1 <- list(c(1,2), list("apple", "orange"), list(1, 2, 3))
typeof(x1)
length(x1)
str(x1)
```

Note that:

- First element of the list `x1` is a numeric vector with length=2
- Second element of the list `x1` is a list with length=2
    - first element is character vector with length=1
    - second element is character vector with length=1
- Third element of the list `x1` is with length=3
    - first element is numeric vector with length=1
    - second element is numeric vector with length=1
    - third element is numeric vector with length=1
    
### You can name each element in the list {-}

```{r}
x2 <- list(a=c(1,2), b=list("apple", "orange"), c=list(1, 2, 3))
str(x2)
```

`names()` function shows names of elements in the list
```{r}
names(x2) # has names
names(x1) # no names
```

### Access individual elements in a "named" list {-}

Syntax: `list_name$element_name`
```{r}
x2 <- list(a=1, b=list("apple", "orange"), c=list(1, 2, 3))
x2$a
typeof(x2$a)
length(x2$a)

typeof(x2$b)
length(x2$b)

typeof(x2$c)
length(x2$c)
```
Note: We'll spend more time practicing "accessing elements of a list" in upcoming weeks

### Compare structure of list to structure of element within a list {-}

```{r}
str(x2)

str(x2$c)
```


### A dataset is just a list! {-}

\medskip A data frame (i.e., R term for a "dataset") is a list with the following characteristics:

- Data type can differ across elements (like all lists)
- Each __element__ in data frame must be a __vector__, not a __list__
    - Each element (column) is a variable
- Each element in a data frame must have the same length
    - The length of an element is the number of observations (rows)
    - so each variable in data frame must have same number of observations
- Each element is named
    - these element names are the variable names

```{r}
load(url("https://github.com/ozanj/rclass/raw/master/data/recruiting/recruit_ps1_small.Rdata"))
```

```{r}
names(df_recruiting)
head(df_recruiting, n=5) # print first 5 rows
```

Additionally, data frames have "attributes"; we'll discuss those in upcoming weeks

### A data frame is a named list {-}



```{r}
df_recruiting
typeof(df_recruiting)
names(df_recruiting)
length(df_recruiting) # length=number of variables
str(df_recruiting)
```
Like any named list, we can examine the elements

- Individual elements of a data frame are the variables
- these variables are vectors with length equal to the number of rows/observations
```{r}
typeof(df_recruiting$state_code)
length(df_recruiting$state_code) # length=number of rows/obs
str(df_recruiting$state_code)
```



### Main takeaways about atomic vectors and lists {-}

Basic data stuctures

1. __(Atomic) vectors__: __logical__, __integer__, __double__, __character__. 
    - each element in vector must have same data type
2. __Lists__: 
    - Data type can differ across elements

Takeaways

- These concepts are difficult; ok to feel confused
- I will reinforce these concepts throughout the course
- Good practice: run simple diagnostics on any new object
    - `length()` : how many __elements__ in the object
    - `typeof()` : what __type__ of data is the object
    - `str()` : hierarchical structure of the object
- These data structures (vectors, lists) and data types (e.g., character, numeric, logical) are the basic building blocks of all object oriented programming languages
- Application to statistical analysis
    - Datasets are just lists
    - The individual elements -- columns/variables -- within a dataset are just vectors
- These structures and data types are foundational for all "data science" applications
    - e.g., mapping, webscraping, network analysis, etc.


### Next Class Session {-}

- No class next Monday 9/5/2022, Labor Day!
- Module 1 Problem Set due on 9/12/2022
  - Rmd file is on course website; you'll need to download it and move it to the "HED696C_RClass/problemsets/module1" folder
  - Practice with knitting and investigating different objects
- Reading for 9/12/2022
  - Wickham & Grolemund _R is for Data Science_
    - Sections 5.1-5.4
  - Xie, Allaire, & Grolemund _R Markdown: The Definitive Guide_
    - Only section 3.1
  - Spend 15 min reviewing this [RMarkdown Reference Guide](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)