---
title: "Module 1 Problem Set"
author: "insert your name here"
date: "insert date here"
output: pdf_document
classoption: landscape

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Welcome to Data Management and Manipulation Using R!

In this problem set, you will practice "knitting," creating objects (i.e., atomic vectors and lists), and investigating those objects.

- Note: Change author to your name & date to current date above in the YAML header (we'll learn more about headers later on...)
- Save this Rmd file to your "HED696C_RClass/problemsets/module1" folder using the naming convention "lastname_module1_ps.Rmd"
- If you're experiencing errors while working on this assignment (or any assignment), start fresh by closing out R (when prompted to "save your workspace" hit "don't save") and re-open the assignment. 

\vspace{30pt}

### Step 1: Practice "knitting" the problem set Rmd file!

- Open your RStudio via the Rproject (`HED96C_RClass.Rproj`) we created in class
- Once in RStudio, in the top menu bar, click on File > Open File...
- Navigate to your HED696C_RClass folder and open your saved `module1_ps.Rmd` file (hint it should be located in: problemsets/module1 folder)
- Open the module1_ps.Rmd file in RStudio
- At the top of the module1_ps.Rmd file, insert your name and the date in the first few lines of this .Rmd file where indicated for you
- Now select the "Knit" tab (icon with blue yarn ball) or use the drop down menu next to the yard ball and select "Knit to PDF"


\vspace{30pt}

### Step 2: Objects in R

**Question 1: What are the two types of vectors in R? How are they different from one another?** 

- ANSWER: TYPE YOUR ANSWER HERE


\vspace{30pt}


**Question 2: How are the two vectors below similar and different?**  

- I have already created two vectors in the R chunk below: `x1` and `x2`
- Run diagnostics of these vectors in the R chunk below using the following functions:
  - `length()`
  - `typeof()`
  - `str()`
- ANSWER: TYPE YOUR ANSWER HERE

```{r}
#object 1
x1 <- c(TRUE, FALSE, FALSE, TRUE)

length(x1)
typeof(x1)
str(x1)

#object 2
x2 <- c(animal="dog", food="kibble", habitat="home", type="domestic")

length(x2)
typeof(x2)
str(x2)
```

\vspace{30pt}


**Question 3**: How are the two objects below similar and different?**  

- I have already created two objects in the R chunk below: `x3` and `x4`
- Run diagnostics of these objects in the R chunk below
- ANSWER: TYPE YOUR ANSWER HERE


```{r}
#object 1
x3 <- list(var1=c(1,2), var2=c("public", "private"), var3=c("AZ", "CA"))
length(x3)
typeof(x3)
str(x3)

#object 2
x4 <- c(1,2,"public", "private", "AZ", "CA")
length(x4)
typeof(x4)
str(x4)

```


\vspace{30pt}



**Question 4**: Access the "var2" element in the `x3` object below. What's the length and type of the var2 element in `x3`? Does the var2 element have a hierarchical structure? 

- I have already re-created the `x3` object in the R chunk below
- Run diagnostics of the "var2" element in the `x3` object in the R chunk below
- ANSWER: TYPE YOUR ANSWER HERE

```{r}
#object 1
x3 <- list(var1=c(1,2), var2=c("public", "private"), var3=c("AZ", "CA"))

length(x3$var2)
typeof(x3$var2)
str(x3$var2)
```

\vspace{30pt}

**Knit to pdf** select the "Knit" tab (icon with blue yarn ball) or scroll down and "Knit to PDF"  
- Submit both .Rmd and pdf
- Use this naming convention "lastname_module1_ps.Rmd"